Liveness-Based Garbage Collection

### 2014-08-30 paper/sections/syntax-and-semantics.tex
The syntax and op. semantics section

\section{The target language---syntax and semantics}
\label{sec:defs}
We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
functions ($\CONS$, $\PRIM$ etc.).
The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
it has eager semantics and restricts programs to be in 
Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
where all actual parameters to functions are variables.
This restriction does not affect expressibility (and indeed we feel free
to ignore it in examples when inessential), but simplifies the analysis formulation.
Additionally, as in the three-address instruction form familiar from compiler texts,
it forces each temporary to be named and function calls to be serialized
(necessary to get an unambiguous definition of liveness).
We further require that each variable in a program is distinct, so that no scope shadowing
occurs---this simplifies proofs of soundness.
In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
or return ($\RETURN$).
The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.

@@

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
% \begin{eqnarray*}
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
\mbox{-- Environment} \\ 
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
  Val}) & \mbox{-- Heap}
% \end{eqnarray*}
\end{array}
\]
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

@@

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \heap, \pp{\pi}{k}
          \rightsquigarrow \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \heap, v_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

@@

\section{Liveness}\label{sec:liveness}

In classical liveness analysis a variable is either `live' (its value
may be used in future computation) or `dead' (definitely not used).
Semantically, a variable is dead at a given program point
if arbitrary changes to its value have no effect on the computation.
Later we will use $\bot$ to represent a value which `explodes' when
it is used in a computation; dead variables can safely have their
value replaced with $\bot$.
For heap-allocated data  we need a richer model of liveness
in that both variables and fields of $\CONS$ cells may be dead or live.
Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
liveness of the structure reachable from a variable is a set of
{\em access paths}
which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
grammar notation.
Thus the liveness of $x$ being
$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
only refer to the second and third members of $x$ considered as a list.
Semantically, access paths are prefix-closed, as accessing a field
requires accessing all the paths from the variable to the field,
and hence the above liveness is properly written 
$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
The classical notions of a scalar variable being live or dead correspond to
$\{\epsilon\}$ and $\{\}$.

The overall liveness (also written {\em liveness environment} for emphasis)
at a program point is conceptually a mapping from
variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
instead of the map
$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
Analogously to classical liveness, the liveness at program point $\pi$
in $\pi:e$ is the liveness just before executing $e$.

A complementary notion to liveness is {\em demand}.  The demand for
expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
context of $e$ may explore of $e$'s result.  So, for example given a
demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
$\pi$ is exactly $x.\sigma$.
The classical analogy of this is in {\em strong liveness}, where
an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
of $x$ at exit from $n$ becomes the demand on $y+z$.
Note that, for an operation like division which
may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
regardless of the liveness of $x$.

We use $\sigma$ to range over demands, $\alpha$ to range over access paths
and $\Lv$ to range over liveness environments. The notation
$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
abuse notation to juxtapose an edge label and a set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
Finally, we use $\Lfonly$ to range over {\em demand transformers}.
These transform demands on a function call
into demands for its formal parameters:
given a user function $f$, defined by
$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.

Note that liveness refers to variables and fields, and not to $\CONS$ cells
(i.e.\ to edges in the memory graph, not to locations themselves).
Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
of location $\ell$.
In the absence of other pointers to heap location $\ell$, we are
certain that the $\CDR$ field of $\ell$ will not be referenced
and may hence be corrupted arbitrarily.
Note therefore, that while  $\ell$ cannot be garbage collected, any
location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
(again provided there are no other aliases to  $\ell$).

%% \comment{
%%   \begin{enumerate}
%% \item Set the framework for liveness. Basic notions: access
%%   paths and expressions.
%% \item An informal working definition of liveness. Minefield will
%%   provide the formal definition.
%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
%%   rules. One of the changes from  the earlier version is the lack of a
%%   post-liveness in  the judgements. This anticipates  the intended use
%%   of  the analysis  for garbage  collection and has  to be  carefully
%%   explained.
%%   \end{enumerate}
%% }

%% We now  set up  the framework for  liveness analysis.  Given  a memory
%% graph, a link in the graph  can be described by starting with a member
%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
%% \CDR\ dereferences to reach the link.  For notational convenience, the
%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
%% dereferences    is    represented   by    a    string   $\alpha    \in
%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
%% in the  memory graph that starts  at $x$ and  traverses adjacent links
%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
%% We call the string $x.\alpha$ an {\em access expression\/}, the string
%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
%% graph   by  the   access  expression   an  {\em   access   path}.   In
%% Figure~\ref{fig:mot-example},        the       access       expression
%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
%% containing the value  $4$.  Most often the memory  graph is clear from
%% the context,  and we shall use  access expressions to  refer to access
%% paths. An  access path  used to refer  to a  link in the  memory graph
%% refers    to     the    last     link    in    the     path.     Thus,
%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
%% set of access paths, then  $x.\sigma$ will represent the set of access
%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
%% \sigma\rbrace$.

%% \subsection{Liveness Analysis}
%% \label{sec:exp-liveness}
%% 
%% Consider the memory graph at a program point $\pi$ during an execution
%% of a program. A  link in the memory graph is said  to be {\em live} at
%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
%% execution  starting from  the  evaluation of  the expression  labelled
%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
%% be live at  $\pi$, if, during {\em any} execution  of the program, the
%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
%% In the first, the path  itself represents the sequence of dereferences
%% to  access  the last  edge,  as,  for  example, while  evaluating  the
%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
%%   implicitly  live} if it is live in the second sense.  Clearly, static
%% determination   of  implicit  liveness   requires  an   analysis  that
%% determines  potential sharing  during  any execution  of the  program.
%% However, the intended application  of our liveness analysis is garbage
%% collection  during which  we have  access  to the  entire heap.   This
%% obviates the need for a  separate sharing analysis and we restrict our
%% attention to explicit liveness only.
%% 
%% 
%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
%% %% \bcdr\ represent application of  the constructor \CONS.  The following
%% %% example motivates the need for these symbols.
%% 
%% %% \begin{example}\label{exmp:motiv-bar}
%% %%   Consider the expression at program point $\pi_1$ in
%% %%   \begin{center}
%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
%% %% 			    $\pi_3\!:${$\cdots$})}}
%% %%   \end{center}
%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
%% %%   \begin{itemize}
%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
%% %%     live at $\pi_1$.
%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
%% %%   \end{itemize}
%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
%% %% \end{example}
%% 
%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
%% %% resulting access patterns may not  be directly useful in that they may
%% %% not describe a  path in the memory graph.  We  define a {\em canonical
%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
%% %% to obtain a canonical access  pattern representing the same edge. This
%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
%% %% cannot  be  used for  nullification  during  garbage collection.   The
%% %% liveness of a such a path is irrelevant, and the path is not considered
%% %% any further.
%% 
%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
%% %% element in $\sigma_2$, i.e.
%% 
%% %% \begin{eqnarray*}
%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
%% %% \end{eqnarray*}

\subsection{Liveness Analysis}
First recall the classical formulation of liveness (as sets of simple variables)
on three-address instructions,
$
\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
     \setminus \mathit{def\/}(I)
     \cup   \mathit{ref\/}(I)
$,
and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
that $\mathit{ref}(I)$ be refined to
$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.

Our liveness analysis formulated in Figure~\ref{fig:live-judge}
is analogous.
% has three analogous aspects.
% \begin{itemize}
% \item
Firstly, the function $\mathit{ref}$, when given a statement $s$,
returns the incremental liveness {\em generated} by $s$.
Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
specifying the demand $\sigma$ on the result of $s$,
to determine which access paths of its free variables are made live.
The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
% The cases for
% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
argument (even if its result is never used).
This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
(note $\sigma$ may be $\{\}$).
In the opposite  sense, the  demand  of $\acar\sigma$  on
(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
that \CONS\ does not, by itself, dereference its arguments.
Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
$\mathit{ref}$
expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
Constants generate no liveness.

% \item
The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
by the latter.
The case $(\LET\ z\leftarrow s\ \IN\ e')$
resembles a three-address instruction:
the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
killing any liveness of $z$ and adding any incremental liveness from $s$.
The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.

% \item
%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
Analogously to classical liveness being computed as
a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
$\Lfonly$ to satisfy the fixed-point property that:
 when we assume $\Lfonly$ to be the family of demand transformers for the program
 then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
 with the assumed  $\Lfonly$.
As usual, there are often multiple solutions to $\Lfonly$; all are safe
(see Section~\ref{sec:correctness})
but we prefer the least one as giving the least liveness subject to safety---and
hence greatest amount of garbage collected.
% \end{itemize}

We make three observations: firstly the rule ({\sc live-define}) has a least solution
as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
({\sc live-define})
resembles the rule for type inference of mutually recursive function definitions,
and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
classically) is due to the functional formulation here.

Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
(representing a fully context-sensitive analysis)
can be safely approximated, for each function, by a {\em procedure summary}
(unifying the contexts in the style of 0-CFA)\@.
The summary consists of a pair of a single demand and, for this demand,
the corresponding tuple of demands the function makes on its arguments.

%%% save two lines to bring up section 3.2.

% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
% enables us to do a context-dependent analysis of a call without having
% to enter the body of the called function.

%% We present  our liveness  analysis using judgement forms $\len$
%% for both expressions and programs.
%% These are given formally in Figure~\ref{fig:live-judge}, but
%% first we an intuitive understanding of the issues involved.
%% Consider the expression:
%% 
%% \begin{center}
%% \begin{uprogram}
%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
%%                         \px) \IN\  
%%            % \UNL{0}
%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
%%                         \pw ) \IN \ldots) 
%% \end{uprogram}
%% \end{center}
%% 
%% 
%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
%% \pw.\epsilon\}$.
%% Informally, this is so since \pw\ and \pz\ are used in
%% (\PRIM~\pz~\pw ).
%% 
%% \begin{enumerate}
%% \item The liveness  at $\pi$ is obtained by  removing the access paths
%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
%%     similar  to liveness  analysis of  imperative programs,  where the
%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
%%     liveness generated and  killed in $s$}. 
%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
%%   as a {\em  demand}.  The demand on an expression consists of  access
%%   patterns  representing all  possible future  uses of  its  value.  A
%%   demand  of $\epsilon$  means that  only the  reference resulting
%% %  from the evaluation  of
%%   from  $(\CAR\,\,\px)$ will be used.
%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
%% \end{enumerate}
%% 
%% In general,  an access path is  live before an  expression because the
%% link represented by it is used (i) while evaluating the expression, or
%% (ii) while using the result of the expression, or (iii) simply because
%% the link  is live at  a later program  point.  When the  expression in
%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
%% the access  path \pw.   Note that \px.\acar\  is live in  this example
%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
%%

% \paragraph{Judgements formally.}
% Let  $\Lv$  denote a  liveness  environment (a function from
% variables to access paths, abused using grammar notation) and
% $\sigma$ be a demand.
% The judgement form $e, \sigma \len \Lv$,
% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
% environment  before $e$  is $\Lv$ and
% is thus analogous to the classical notion of {\em transfer function}.

%    It is  interesting to  note that,
% unlike traditional liveness analysis,  our judgements do not require a
% liveness after the expression. This  is because of the coming together
% of  two reasons.   First, it  will be  clear while  discussing garbage
% collection,  that it is  enough for  our purposes  to assume  that the
% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
% restriction of our  programs to ANF ensures that  the post-liveness of
% each  expression  $e$  inside   the  function  body  continues  to  be
% $\emptyset$. Thus the post-liveness can be dropped altogether.

%==============================================================
\begin{figure}[t]  
% \footnotesize
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
          &=& \{x.\sigma\} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=& \{x.\epsilon,y.\epsilon\} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \\
\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
        \{x.\epsilon\} \\
\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
        \Lv
           \setminus \{x.*\}
           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
\end{eqnarray*}

\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}

\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%==============================================================

% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
% liveness generated  before the right-hand-side of  a \LET\ definition.
% Constants    do   not   generate    any   liveness.     The   operator
% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
% that \CONS\ does not, by itself, dereference its arguments.
% 
% 
% Unlike primitives, the mapping from the demand on a user-function call
% to the  demands on its arguments  has to be derived  from the function
% body itself---as a fixed point for a recursive function.
% The function  \Lfonly\ captures this mapping for
% all functions  defined in the  program.  For a  function $\mathit{f}$,
% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
% a    call     to    the     $i$th    argument.     The     rule
% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
% confirmed by the body of all the functions, then the guess is correct.
% {\color{blue}Note the similarity of this rule to type-inference
% for recursive program schemes.}

% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
% liveness generated before $s$.

>>>>

\section{The target language---syntax and semantics}
\label{sec:defs}
We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
$p$ over  primitive functions ($\CONS$, $\PRIM$ etc.).   The syntax of
our language  is shown  in Figure~\ref{fig:lang-syntax}; it  has eager
semantics and  restricts programs to be in  Administrative Normal Form
(ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
functions   are   variables.   This   restriction   does  not   affect
expressibility (and indeed we feel  free to ignore it in examples when
inessential). In addition, this  form also explicates the formation of
closures.  We further  require  that  each variable  in  a program  is
distinct, so  that no scope shadowing  occurs---this simplifies proofs
of soundness.  In this  formulation expressions: either perform a test
($\SIF$), make a computation step ($\LET$) or return ($\RETURN$).  The
$\RETURN$ keyword is logically redundant, but we find it clarifies the
semantics and analysis.

@@

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
  Val}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
\mathit{Clo}) & \mbox{-- Heap}
\end{array}
\]
%Note that in the eager case, if a cons had a numeric argument, this
%fact was known when the cons was created. Thus you could have unboxed
%numbers. This is not possible now. For instance, the argument of a
%cons could be a closure which evaluates to a number.
%Decision - numbers are boxed
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

@@

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \stk, \heap,   \pp{\pi}{k}
          \rightsquigarrow \rho, \stk, \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
%%\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \ell \bullet \stk,\heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \rho, \stk, \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (e_1, e_2)}
        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \rho, \stk, \heap, e_1}
\end{minipage}
\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = e}
        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \rho, (\CAR \bullet):\stk, \heap, v_1}
\end{minipage}\\[0.5ex]

\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

@@
